Agda Meets Accelerate

### 2013-03-05 paper/main.tex
second round of revisions

The general type for \verb+use+ is
%
\begin{hcode}
use :: Elt e => Array sh e -> Acc (Array sh e)
\end{hcode}
%
where the class \verb+Elt+ characterizes all types that may be held in Accelerate arrays. These are currently primitive types and tuples.

Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are
predefined, but to enable shape polymorphic computations, along the
lines pioneered in the Haskell array library
Repa~\cite{keller-etal:repa}, shapes are inductively defined using
type-level snoc lists built from the data types \texttt{Z} and \texttt{:.}:
%
\begin{hcode}
data Z       = Z
data sh :. i = sh :. i
\end{hcode}
%
Hence, the definitions of the dimensions:
%
\begin{hcode}
type DIM0 = Z
type DIM1 = DIM0 :. Int
-- and so on
\end{hcode}

@@

To demonstrate the Agda embedding in action, we translate the
dot product example from Section~\ref{sec:accelerate} to Agda.
\begin{code}
dotp : forall {E : Elt} {{p : Numeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{code}
Unlike the Accelerate code, this function is polymorphic with respect
to the array element type, provided it is numeric.\footnote{In Agda,
  arguments in double curly braces are \emph{instance
    arguments}~\cite{DevriesePiessens2011} that are aggressively
  inferred. We use them like type class constraints in Haskell.}
The length parameter $n$ ensures that the two input
vectors have the same size. The \texttt{PreVector} type of the
arguments corresponds to the plain \texttt{Vector} type in Accelerate,
whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
  (Scalar E)}---a piece of abstract syntax.

@@

Some parallel reduction operations require their base operation to be
associative to return a predictable result. Here are two examples from
Accelerate. 
\begin{hcode}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{hcode}
In both cases, the text of the documentation says that ``the first
argument needs to be associative'' and furthermore the \texttt{fold1}
documentation ``requires the reduced array to be non-empty''.
The second requirement can be enforced by asking for a suitable proof
object on each call of \texttt{fold1}:
\begin{verbatim}
fold1 : ... -> Array (sh :< n >) E -> (size sh * n > 0)
            -> Array sh E
\end{verbatim}
The first requirement can be rephrased to saying that the first two
parameters of \texttt{fold} together form a monoid, which requires an
associative operation with a unit element. The concept of a monoid
can be formalized in Agda, which has indeed been done in the standard
library. Unfortunately, the formalization from the library cannot be
used because Accelerate deals with ASTs, not with values. So, a
formalization is required that states that the meaning of an
AST-encoded function is associative and the meaning of another
AST-encoded constant is its unit element. Given that Accelerate
encodes AST construction using higher-order abstract syntax, such a
formalization is not straightforward. Moreover, even given expressions
with a fixed meaning, there is no general shape for associative
functions, so that proofs can only be done for special cases.

@@

Some readers may object that neither addition nor multiplication of
floating point numbers is associative
\cite{DBLP:journals/csur/Goldberg91}. However, for advanced
optimizations, the exploitation of algebraic laws is a necessity and
the involved degradation of precision or change of result is accepted
or account for in the error estimates. Moreover, there are
other operations, like min or max, that are commonly used with
\texttt{fold}-like operations, which are truly associative. Last, but
not least, the associativity declarations serve as an important
documentation that passing an inherently non-associative function will
produce unpredictable, implementation-dependent results.

@@

Because of Agda's limited support for numeric data types,
we embed numeric literals for integers and floating point numbers
using a string with an explicit type annotation that determines the
parsing of the string. Here are some example embeddings:
\begin{verbatim}
"3.1415926" ::: Float
"6.0221415E23" ::: Double
\end{verbatim}
Recall that \texttt{Float} and \texttt{Double} are not
types, but rather values of type \texttt{Elt}.
The \texttt{:::} operation is the workhorse of the embedding:
\begin{verbatim}
_:::_ : (s : String) -> (E : Elt) 
     -> {{nu : Numeric E}} -> {p : T (s parsesAs E)} -> Exp E
s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
\end{verbatim}
The arguments \texttt{s} and \texttt{E} are explicit, but the
remaining ones are inferred by Agda.
As mentioned before, the argument \texttt{nu} is an instance argument; it is automatically
filled-in with a suitably typed value that is currently in scope
\cite{DevriesePiessens2011}. Here, the predicate 
\texttt{Numeric} plays the role of a type class that characterizes
the numeric types.

@@

>>>>

In the general type for \verb+use+, the class \verb+Elt+ characterizes all types that may be held in Accelerate arrays. These are currently primitive types and tuples.
%
\begin{hcode}
use :: Elt e => Array sh e -> Acc (Array sh e)
\end{hcode}
%
Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are
predefined, but to enable shape polymorphic computations, along the
lines pioneered in the Haskell array library
Repa~\cite{keller-etal:repa}, shapes are inductively defined using
type-level snoc lists built from the data types \texttt{Z} and
\texttt{:.}. The use of snoc lists simplifies the type signatures of
fold operations that reduce or abstract over the least significant dimensions.
%
\begin{hcode}
data Z       = Z
data sh :. i = sh :. i
-- Types for often used dimensions
type DIM0 = Z
type DIM1 = DIM0 :. Int
-- and so on
\end{hcode}

@@

To demonstrate the Agda embedding in action, we translate the
dot product example from Section~\ref{sec:accelerate} to Agda.\footnote{In Agda,
  arguments in double curly braces are \emph{instance
    arguments}~\cite{DevriesePiessens2011} that are aggressively
  inferred. We use them like type class constraints in Haskell.}
\begin{code}
dotp : forall {E : Elt} {{p : Numeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{code}
Unlike the Accelerate code, this function is polymorphic with respect
to the array element type, provided it is numeric.
The length parameter $n$ ensures that the two input
vectors have the same size. The \texttt{PreVector} type of the
arguments corresponds to the plain \texttt{Vector} type in Accelerate,
whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
  (Scalar E)}---a piece of abstract syntax.

@@

Some parallel reduction operations require their base operation to be
associative to return a predictable result. Here are two examples from
Accelerate. 
\begin{hcode}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{hcode}
In both cases, the text of the documentation says that ``the first
argument needs to be associative'' and furthermore the \texttt{fold1}
documentation ``requires the reduced array to be non-empty''.
The second requirement can be enforced by asking for a suitable proof
object on each call of \texttt{fold1}:
\begin{verbatim}
fold1 : ... -> Array (sh :< n >) E -> (size sh * n > 0)
            -> Array sh E
\end{verbatim}
The first requirement can be rephrased to saying that the first two
parameters of \texttt{fold} together form a monoid, which requires an
associative operation with a unit element. The concept of a monoid
can be formalized in Agda, which has indeed been done in the standard
library. Unfortunately, the formalization from the library cannot be
used because Accelerate deals with ASTs, not with values. So, a
formalization is required that states that the meaning of an
AST-encoded function is associative and the meaning of another
AST-encoded constant is its unit element. Given that Accelerate
encodes AST construction using higher-order abstract syntax, such a
formalization is not straightforward. Moreover, even given expressions
with a fixed meaning, associativity has to be proved on a case by case
basis.

@@

Some readers may object that neither addition nor multiplication of
floating point numbers is associative
\cite{DBLP:journals/csur/Goldberg91}. However, for advanced
optimizations, the exploitation of algebraic laws is a necessity and
the involved degradation of precision or change of result is accepted
or accounted for in the error estimates. Moreover, there are
other operations, like min or max, that are commonly used with
\texttt{fold}-like operations, which are truly associative. Last, but
not least, the associativity declarations serve as important
documentation that passing an inherently non-associative function will
produce unpredictable, implementation-dependent results.

@@

Because of Agda's lack of support for overloaded numeric literals,
we embed numeric literals for integers and floating point numbers
using a string with an explicit type annotation that determines the
parsing of the string. Here are some example embeddings:
\begin{verbatim}
"3.1415926" ::: Float
"6.0221415E23" ::: Double
\end{verbatim}
Recall that \texttt{Float} and \texttt{Double} are not
types, but rather values of type \texttt{Elt}.
The \texttt{:::} operation is the workhorse of the embedding:
\begin{verbatim}
_:::_ : (s : String) -> (E : Elt) 
     -> {{nu : Numeric E}} -> {p : T (s parsesAs E)} -> Exp E
s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
\end{verbatim}
The arguments \texttt{s} and \texttt{E} are explicit, but the
remaining ones are inferred by Agda.
As mentioned before, the argument \texttt{nu} is an instance argument; it is automatically
filled-in with a suitably typed value in scope
\cite{DevriesePiessens2011}. As before, the predicate 
\texttt{Numeric} plays the role of a type class that characterizes
the numeric types.

@@

This setup results in a flexible way of handling literals. It has
worked well in our examples.
